Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-12-01 tmm.tex
Polish the first two sections

In video streaming, the network condition significantly influences the user's watching experience. To provide better streaming service, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of adaptive video manycast \cite{Xue15}. Although these studies offer a promising way to optimize network conditions, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to cellular network). Even in the same network, the user's bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams, each with different video quality, are stored at the server, and the client selects one of them according to currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, an extension of the H.264/AVC video coding standard \cite{SVCOverview}, which allows scalability of video bitrate and quality with only one single bitstream. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate data rate and video quality as required.

Due to the easy deployment of simulcast systems \cite{Bouten14}, they are already used by many commercial video websites (e.g., YouTube \cite{YouTube}) to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams require a lot of computation time and disk space. In order to provide seamless quality adaptation at reasonable time and space cost, the adoption of SVC is necessary. An SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), and performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast. Therefore, to adopt SVC in the video streaming framework is technically a better option and has attracted much attention in the study \cite{Chuah12}-\cite{Cicalo14}.

There are two key problems in the SVC adaptive video streaming system that need to be primarily addressed. First, as the prerequisite of the adaption to bandwidth change, the whole bitstream needs to be extracted so that sub-streams with various bitrates can be obtained. This is called the bitstream extraction problem. Optimal bitstream extraction algorithm aims to extract a sub-stream with best video quality, or minimal distortion, under the constraint of a given bitrate. Second, when the bandwidth changes with time, it's necessary to make decisions about whether and how to adjust the bitrate or video quality. This problem is called quality control, which means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. A good quality control algorithm should determine a suitable bitrate for current bandwidth condition, and keep the video quality both smooth and at a high level.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to enable better quality of service for the user. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective Linear Error Model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, on the basis of the distortion estimation enabled by LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its Rate-Distortion (R-D) impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves much lower quality fluctuation with higher quality level.
\end{itemize}

@@

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related works on them. Then in section \ref{sec:bitstream-extraction}, the linear error model is proposed, and the solution for the bitstream extraction problem based on the model is present. After that, the PID-based quality control algorithm is proposed in section \ref{sec:quality-control}. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper's work and point out its promising applications and extensions in the future.

@@

In this section, we analyze the two most concerned problems in SVC adaptive video streaming, i.e., bitstream extraction and quality control, and introduce some related works on them.

@@

Bitstream extraction is the operation of choosing a subset of data packets from a whole SVC bitstream to form a sub-stream. It makes the bitrate adjustable and is the prerequisite of an adaptive streaming system. Fig. \ref{fig:Bitstream-Extraction} illustrates the structure of an SVC stream and a possible way to extract a sub-stream from it. The data packets in the stream are divided into different layers. The temporal layer (T0$\sim$T3) reflects the reference or dependency relationship between the frames. The arrows on the top point from the referenced frame to the frame which is dependent on it. The quality layer (Q0$\sim$Q2) indicates scalability of video quality within each frame. Data packets in the higher quality layers (i.e., Q1 and Q2) are partially discarded to implement the extraction. For example, the dash line demonstrates the extraction which will keep the data packets below the line and discard all the packets above it.

@@

As we can see, in its nature SVC bitstream extraction is a combinatorial optimization problem. There are many combinations of packets to be discarded, and we want to choose a combination that maximizes the video quality of the extracted sub-stream, under the constraint of a given bitrate. More specifically, this problem fits in the form of ``0-1 Knapsack Problem" \cite{Knapsack}. Each data packet can be treated as an object with certain weight (the packet's size) and value (its contribution to the video quality). The solving process is mainly to decide whether or not a packet should be included in the final extracted sub-stream, or in the term of ``0-1 Knapsack Problem", to choose which object to pack up.

For a typical ``0-1 Knapsack Problem", the optimal solution can generally be found using the dynamic programming algorithm. However, this is not operational for the bitstream extraction problem, mainly due to the dependence between the stream's data packets. First, the subset of packets to be extracted cannot be chosen arbitrarily, since if a packet is included, all the packets it depends on must also be included. For example, in Fig. \ref{fig:Bitstream-Extraction}, we cannot include the Q2 packet of a frame without including the Q1 packet of the same frame. Second, the value of a packet is not obvious and certain, since a packet's contribution to video quality may depend on other packets. Unless we discard a packet and calculate the distortion it causes for real, there is no way to exactly tell its value. In fact, to the best of our knowledge, no method presented in the literature claims to find the optimal solution of the SVC bitstream extraction problem. All the effort is made to obtain a solution that is as good as possible.

In \cite{Amonou07}, Amonou et al. proposed a bitstream extraction framework based on the concept of \textit{Quality Layers}. This framework has been adopted by the SVC reference software, i.e., JSVM  \cite{JSVM}, and achieves better R-D performance compared with the basic extractor of JSVM. In Amonou's framework, a Quality Layer value, which reflects a packet's R-D impact, is calculated and used for R-D optimized bitstream extraction. The process of obtaining distortion impact of every packet is extremely time consuming, and the distortion estimation accuracy can also be further improved. So other distortion/error models were proposed to either reduce the computational complexity or improve the estimation accuracy. Sun et al. \cite{Sun09} and Maani et al. \cite{Maani09} constructed models to calculate distortion based on the drift propagation between frames. Sun et al.'s method is almost as good as JSVM in regard of performance, but has much reduced complexity. Maani et al.'s model can achieve better estimation accuracy than JSVM; however, since it is a training-based model, more computation is needed to get robust model parameters. Ramanathan et al. \cite{Ramanathan12} proposed their concept and calculation method of the ``Quality Metadata'' to signal and use in bitstream extraction. Yang et al. \cite{Yang13} proposed a method based-on simulated annealing to determine and assign priority values used in bitstream extraction. In both methods, the accuracy of distortion estimation is the key of the performance.

In our previous work \cite{Zhang12}, we noticed the mismatch between quadratic calculations in distortion criteria like MSE or PSNR and the essential linear operations in the coding process, and proposed a linear error model to directly estimate pixel value errors before MSE or PSNR is calculated. The linear error model exploits the linear feature of pixel value errors and achieves good performance in distortion estimation and bitstream extraction with reference. However, this work assumes that the original video sequence is available for reference, which is not always true. Adaptation to handle the special characteristic of the no-reference case is lacking in the previous work and included in this paper.

@@

Since the next generation video coding standard High Efficiency Video Coding (HEVC) \cite{HEVC} adopts the similar coding framework as its predecessor H.264/AVC, we expect to apply this paper's model and approach of bitstream extraction to the emerging HEVC standard and its scalable extension. As for the PID-based quality control algorithm, it is only network-concerned thus should be easily integrated into other streaming systems, e.g., simulcast system like DASH.

>>>>

In video streaming, the network condition significantly influences the user's watching experience. To provide better network condition, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm named OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration
of OpenFlow-controlled network orchestration for video manycast \cite{Xue15}. Although these studies have offered a promising alternative, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to Cellular). Even in the same network, the end-to-end bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable, and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams, each with different video quality, are stored at the server, and the client selects one of them according to currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, an extension of the H.264/AVC video coding standard \cite{SVCOverview}, which allows scalability of video quality and bitrate with only one single bitstream. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate video quality and bitrate as required.

Due to the easy deployment of simulcast systems \cite{Bouten14}, they are already used by many commercial video websites (e.g., YouTube \cite{YouTube}) to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams require a lot of computation time and disk space. In order to provide a wide range of seamless quality adaptation at reasonable time and space cost, the adoption of SVC is necessary. An SVC video bitstream can be extracted at the fine granularity of one NALU (a very small partition of data), and performance analysis \cite{SVCPerformance} shows that SVC clearly outperforms simulcast coding in terms of compression efficiency. Therefore, to adopt SVC in the video streaming framework is technically a better option and has attracted much attention in the study \cite{Chuah12}-\cite{Cicalo14}.

While these studies mostly focus on Quality of Service (QoS) optimization and resource allocation for specific networks, there are two key problems that need to be primarily addressed when adopting SVC in adaptive video streaming. First, as the prerequisite of bitrate adaption, the whole bitstream needs to be extracted so that sub-streams with various bitrates can be obtained. This is called the bitstream extraction problem. Optimal bitstream extraction algorithm aims to extract a sub-stream with best video quality, or minimal distortion, under the constraint of a given bitrate. Second, when the bandwidth changes with time, it is necessary to make decisions about whether and how to adjust the video quality or bitrate. This problem is called quality control, which means controlling the video quality (or the bitrate, from the transmission point of view) to follow the change of the bandwidth. A good quality control algorithm should determine a suitable bitrate according to current network condition, and keep the video quality both smooth and at a high level.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to enable better quality of service for the user. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective Linear Error Model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, utilizing distortion estimation with LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its Rate-Distortion (R-D) impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) controller is proposed to replace the native one in DSS, and much lower quality fluctuation is achieved.
\end{itemize}

@@

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related work on them. Then in section \ref{sec:bitstream-extraction}, the linear error model and the solution for the bitstream extraction problem are present. After that, the PID-based quality control algorithm is proposed in section \ref{sec:quality-control}. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper's work and point out its promising applications and extensions in the future.

@@

In this section, we analyze the two most concerned problems in SVC adaptive video streaming, i.e., bitstream extraction and quality control, and introduce some related work on them.

@@

Bitstream extraction is the operation of choosing a subset of data packets from a whole SVC bitstream to form a sub-stream. Fig. \ref{fig:Bitstream-Extraction} illustrates the structure of an SVC bitstream and a possible way to extract a sub-stream from it. The data packets in the bitstream are divided into different layers. The temporal layer (T0$\sim$T3) reflects the reference or dependency relationship between the frames. The arrows on the top point from the referenced frame to the frame which is dependent on it. The quality layer (Q0$\sim$Q2) indicates scalability of video quality within each frame. Data packets in the higher quality layers (i.e., Q1 and Q2) are partially discarded to implement the extraction. For example, the dash line demonstrates the extraction which will keep the data packets below the line and discard all the packets above it.

@@

As we can see, in its nature SVC bitstream extraction is a combinatorial optimization problem. There are many combinations of packets to be discarded, and we want to choose a combination that maximizes the video quality of the extracted sub-stream, under the constraint of a given bitrate. More specifically, this problem is in the form of ``0-1 Knapsack Problem" \cite{Knapsack}. Each data packet can be treated as an object with certain weight (the packet's size) and value (its contribution to the video quality). The solving process is mainly to decide whether or not a packet should be included in the final extracted sub-stream, or in the term of ``0-1 Knapsack Problem", to choose which object to pack up.

For a typical ``0-1 Knapsack Problem", the optimal solution can generally be found using the dynamic programming algorithm. However, this is not operational for the bitstream extraction problem, mainly due to the dependence between the data packets. First, the subset of packets to be extracted cannot be chosen arbitrarily, since if a packet is included, all the packets it depends on must also be included. For example, in Fig. \ref{fig:Bitstream-Extraction}, we cannot include the Q2 packet of a frame without including the Q1 packet of the same frame. Second, the value of a packet is not obvious and certain, since a packet's contribution to video quality depends on other packets and may vary in different extractions. There is no way to exactly tell the video quality of an extracted sub-stream, unless we perform the extraction and calculate the distortion for real. In fact, to the best of our knowledge, no method presented in the literature claims to find the optimal solution of the SVC bitstream extraction problem. All the effort is made to obtain a solution that is as good as possible.

In \cite{Amonou07}, Amonou et al. proposed a bitstream extraction framework based on the concept of \textit{Quality Layers}. This framework has been adopted by the SVC reference software, i.e., JSVM \cite{JSVM}, and achieves better R-D performance compared with the basic extractor of JSVM. In Amonou's framework, a Quality Layer value, which reflects a packet's R-D impact, is calculated and used for R-D optimized bitstream extraction. The process of obtaining distortion impact of every packet is extremely time consuming, and the distortion estimation accuracy can also be further improved. So other distortion/error models were proposed to either reduce the computational complexity or improve the estimation accuracy. Sun et al. \cite{Sun09} and Maani et al. \cite{Maani09} constructed models to calculate distortion based on the drift propagation between frames. Sun et al.'s method is almost as good as JSVM in regard of performance, but has much reduced complexity. Maani et al.'s model can achieve better estimation accuracy than JSVM; however, since it is a training-based model, more computation is needed to get robust model parameters. 
With relatively simple distortion estimation, Yang et al. \cite{Yang13} proposed a method based-on simulated annealing to determine priority values used in bitstream extraction. Ramanathan et al. \cite{Ramanathan12} proposed their concept and calculation method of the ``Quality Metadata'' to signal and use in bitstream extraction. In both methods, the accuracy of distortion estimation still affects the performance a lot.

In our previous work \cite{Zhang12}, we proposed a linear error model to directly estimate pixel value errors before MSE or PSNR is calculated, which achieves good performance in distortion estimation and bitstream extraction. However, this work assumes that the original video sequence is available for reference, which is not always true. How to handle the special characteristic of the no-reference case is lacking in the previous work and included in this paper.

@@

Since the next generation video coding standard High Efficiency Video Coding (HEVC) \cite{HEVC} adopts the similar coding framework as its predecessor H.264/AVC, we expect to apply this paper's model and approach of bitstream extraction to the emerging HEVC standard and its scalable extension. As for the PID-based quality control algorithm, it is only network-concerned thus should be easily integrated into other streaming systems, e.g., simulcast system like DASH. Another direction of future work is to carry out some subjective tests, comparing the performance of the proposed system from the perceived quality point of view.
